---
title: "nlmr_simulations_confounder"
author: "Fergus Hamilton + Tianyuan Lu"
date: "2023-09-06"
output: html_document
---

Simulation with confounder was conducted with the following setting, where there was no causal effect of the exposure $X$ on the outcome $Y$ such that the association between $X$ and $Y$ was purely due to the confounder $U$. The instrument $G$ and the confounder $U$ can have an interaction effect on $X$.

$X=0.3G+\beta_{UX}U+\beta_{GUX}GU+\epsilon_X$

$Y=\beta_{UY}U+\epsilon_Y$

We explored the impact of the $G$-by-$U$ interaction effect size by fixing $\beta_{UX}=0.3$ and $\beta_{UY}=0.3$, and varying $\beta_{GUX}\in\{-0.1,-0.02,0,0.02,0.1\}$. When $\beta_{GUX}=0$, there was no $G$-by-$U$ interaction effect.

We explored the impact of the confounding effect size by fixing $\beta_{GUX}=-0.1$, and separately varying $\beta_{UY}\in\{0,0.05,0.3\}$ and $\beta_{UX}\in\{0,0.05,0.3\}$. When $\beta_{UY}=0$, there was no confounding effect, despite the $G$-by-$U$ interaction effect on $X$.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## Load needed packages ###
library(tidyverse)
library(broom)
library(patchwork)
library(mgcv)
library(metafor)
library(faux)
library(lmtest)
library(vroom)

## Load my functions ###

# YOU'll have to change this!
f = list.files("required_functions/")

f
for(file in f){
  print(file)
  source(paste0("required_functions/",file))
}

```


```{r}
analyses <- tibble(
  exposures = c("x"),
  outcomes = c("y")
)

# FIGURE 1 of this section
map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 2023, reps = 50, bux = 0.3, buy = 0.3, bgux = -0.1, .f = generate_outcome_plots_linear_replicates_confounder, covariates = "u_incomplete", u_cor =0.1)

# SUP FIGURE 1 of this section
map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 2023, reps = 100, bux = 0.3, buy = 0.3, bgux = -0.02, .f = generate_outcome_plots_linear_replicates_confounder)

# SUP FIGURE 2
map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 2023, reps = 100, bux = 0.3, buy = 0.3, bgux = 0, .f = generate_outcome_plots_linear_replicates_confounder)

# SUP FIGURE 3
map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 2023, reps = 100, bux = 0.3, buy = 0.3, bgux = 0.02, .f = generate_outcome_plots_linear_replicates_confounder)

# SUP FIGURE 4
map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 2023, reps = 100, bux = 0.3, buy = 0.3, bgux = 0.1, .f = generate_outcome_plots_linear_replicates_confounder)

#  sup figure 5 and 6 - varying byz
map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 2023, reps = 100, bux = 0.05, buy = 0.3, bgux = -0.1, .f = generate_outcome_plots_linear_replicates_confounder)

#  sup figure 6

map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 2023, reps = 100, bux = 0, buy = 0.3, bgux = -0.1, .f = generate_outcome_plots_linear_replicates_confounder)


# Figure 2 of this section
map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 2023, reps = 100, bux = 0.3, buy = 0.05, bgux = -0.1, .f = generate_outcome_plots_linear_replicates_confounder)


#  sup figure 6

map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 2023, reps = 100, bux = 0.3, buy = 0, bgux = -0.1, .f = generate_outcome_plots_linear_replicates_confounder)




```

confirm the IV and confounder become associated

```{r}

s <- sim_mydata_outcomes_confounder(bux = 0.05, buy = 0.3, bgux = -0.1, seed = 2023)

s %>%
  mutate(strat = generate_ranked_strata(g,x,10)) %>%
  ggplot(aes(x = x, y = u)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ strat)
```

Kates suggestion: add a strong confounder of x and y (v), and interaction (bgux = 0.01)

```{r}
map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 2023, reps = 100, bux = -0.3, buy = 0, bgux = -0.1, bv= 1,.f = generate_outcome_plots_linear_replicates_confounder)

map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 2023, reps = 100, bux = -0.3, buy = 0, bgux = -0.1, bv= 0.3,.f = generate_outcome_plots_linear_replicates_confounder)

map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 2023, reps = 100, bux = 0.3, buy = 0, bgux = -0.1, bv= 1,.f = generate_outcome_plots_linear_replicates_confounder)


map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 2023, reps = 100, bux = 0, buy = 0, bgux = -0.1, bv= 1,.f = generate_outcome_plots_linear_replicates_confounder)
```

Test adjusting for the interacting variable

```{r}
map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 2023, reps = 100, bux = 0, buy = 0, bgux = -0.1, bv= 1,.f = generate_outcome_plots_linear_replicates_confounder, covariates = "v")
```


Now, fernandos suggestion of a variance approach

```{r}
map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 203, reps = 100,
     bux= 0, buy = 0, g_sd = 0.5, bx = 0.3,.f = generate_outcome_plots_vQTL)

map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 203, reps = 100,
     bux= 0.5, buy = 0.5, g_sd = 0.5, bx = 0.3,.f = generate_outcome_plots_vQTL)

#  test no mean effect
map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 23, reps = 100,
     bux= 0.5, buy = 0.5, g_sd = 0.5, bx = 0,.f = generate_outcome_plots_vQTL)


map2(.x = analyses$exposures, .y = analyses$outcomes, seed = 23, reps = 100,
     bux= 1, buy = 1, g_sd = 0, bx = 0.3, u_sd = 1,.f = generate_outcome_plots_vQTL)
```

Now to explain why

```{r}
s <- sim_mydata_outcomes_vQTL(seed = 203, n = 100000, g_sd = 1, bx = 0)

s %>%
  mutate(strat = generate_ranked_strata(g,x,10)) %>%
  ggplot(aes(x = g, y = y)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ strat)




p1 <- s %>%
  as_tibble() %>%
  mutate(pre_strata = ntile(g, 1000)) %>%
  group_by(pre_strata) %>%
  summarise(sd_x = sd(x)^2) %>%
  ggplot() +
  geom_point(aes(x = pre_strata, y = sd_x)  ) +
  theme_bw() +
  xlab("Pre-strata") +
  ylab("Variance of the exposure")

p2 <- s %>%
  as_tibble() %>%
  mutate(pre_strata = ntile(g, 1000)) %>%
  mutate(dr = generate_ranked_strata(g,x,100)) %>%
  ggplot() +
  geom_point(aes(x = g, y = x), alpha = 0.1) +
    geom_smooth(aes(x = g, y = x))  +
  theme_bw() +
  xlab("Level of the IV (g)") +
  ylab("Level of the exposure (x)")



p3 <- s %>%
  as_tibble() %>%
  mutate(pre_strata = ntile(g, 1000)) %>%
  mutate(dr = generate_ranked_strata(g,x,100)) %>%
  filter(dr == 100 | dr == 1) %>%
  # filter(strat == 1000 | strat == 1) %>%
  ggplot() +
  geom_smooth(aes(x = g, y = x, group = dr))  +
  geom_point(aes(x = g, y = x, colour = pre_strata)) +
  theme_bw() +
  xlab("Level of the IV (g)") +
  ylab("Level of the exposure (x)")





p1/p2/p3 + plot_annotation(tag_levels = "A")
ggsave("figures/explain_why_variance_matters.png", height = 12, width = 8)

  
```

